---
title: "Fast Food"
description: |
  Analysis of Fast Food Calorie data from TidyTuesday
author: Josh Goldhaber
date: September 24, 2025
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)
library(tidyverse)
```

```{r}
#| fig-alt: "Bar chart displaying the top 10 fast food chains ranked by average calories per menu item. Each bar represents a restaurant, with the x-axis showing the restaurant names and the y-axis showing the average calories. The chart highlights calorie differences across chains."

fastfood <- read_csv("data/fastfood_calories.csv", show_col_types = FALSE)

avg_cal <- fastfood %>%
  group_by(restaurant) %>%
  summarise(mean_calories = mean(calories, na.rm = TRUE),
            n_items = n()) %>%
  arrange(desc(mean_calories))

avg_cal %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(restaurant, mean_calories),
             y = mean_calories)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(
    title = "Top 10 Fast Food Chains by Average Calories per Menu Item",
    x = "Restaurant",
    y = "Average Calories"
  ) +
  theme_minimal()

```

## Analysis:
This chart shows the fast-food chains with the highest average calories per menu item, highlighting meaningful differences across major brands. McDonald’s, Sonic, and Burger King top the list, each averaging over 600 calories per item—indicating that their menus tend to feature larger portions or more calorie-dense foods.

Restaurants like Arby’s, Dairy Queen, and Subway fall in the mid-range (roughly 520–560 calories), suggesting a mix of lighter and heavier options. At the lower end of the top ten, Taco Bell and Chick-fil-A average under 450 calories, indicating comparatively lighter items on average.

Overall, the data shows substantial variation across chains, with some offering consistently higher-calorie options. This may reflect differences in menu composition (e.g., fried foods, desserts, portion sizes), target customers, or brand positioning within the fast-food market.


## Source:
Hughes, E. (2018). Fast food entrée data [Data set]. TidyTuesday, rfordatascience.
Retrieved from https://github.com/rfordatascience/tidytuesday/blob/cd0df1831dc506ea0a6f1295d7605142120d6dbd/data/2018/2018-09-04/fastfood_calories.csv