---
title: "Recap: Final Presentation"
author: "Josh Goldhaber"
subtitle: "December 10, 2025 "
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---   

# Overview 

- **Text Analysis:**  
  Understanding thematic patterns and linguistic structure in Netflix titles and descriptions.

- **Permutation Tests: Silver vs. Gold Volatility:**  
  A permutation test examining whether silver is more volatile than gold from August–October.

- **Traffic Stops with SQL:**  
  Combining multi-city policing datasets, harmonizing outcomes, and analyzing citation/warning disparities.

# Text Analysis Recap
- Used **regular expressions** to break down text into words.
- Removed **stop words** to focus on meaningful terms.
- Analyzed **word frequency** to identify common themes.

## Text Analysis Code

### Package Installation
```{r}
# Install required packages
install.packages(c("tidyverse", "stringr", "tidytext", "ggplot2"))
```

## Setup
```{r}
# Load libraries and set chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)
library(tidyverse)
library(stringr)
library(tidytext)
library(ggplot2)
```

## Proportion of Netflix Titles Containing Numbers
```{r}
#| eval: true
netflix_titles <- netflix_titles |>
  mutate(has_number = str_detect(title, "\\d+"))

table(netflix_titles$has_number)
prop.table(table(netflix_titles$has_number))
```

## Counting Thematic Keywords in Descriptions
```{r}
#| fig-alt: "Bar chart showing total mentions of the words 'love', 'death', and 'mystery' across all Netflix descriptions. Each bar represents the overall frequency of each theme, allowing comparison of how often these motifs appear in Netflix title summaries."
#| eval: true
netflix_titles <- netflix_titles |>
  mutate(
    love_mentions    = str_count(description, "\\b(?i)love\\b"),
    death_mentions   = str_count(description, "\\b(?i)death\\b"),
    mystery_mentions = str_count(description, "\\b(?i)mystery\\b")
  )

theme_counts <- netflix_titles |>
  summarise(
    love = sum(love_mentions, na.rm = TRUE),
    death = sum(death_mentions, na.rm = TRUE),
    mystery = sum(mystery_mentions, na.rm = TRUE)
  ) |>
  pivot_longer(cols = everything(), names_to = "theme", values_to = "count") 

ggplot(theme_counts, aes(x = theme, y = count, fill = theme)) +
  geom_col() +
  labs(
    title = "Common Themes in Netflix Descriptions",
    x = "Theme",
    y = "Count of Mentions",
    fill = "Theme"
  ) +
  theme_minimal()
```

### Finding Adjectives Before "Story"
```{r}
#| eval: true
story_words <- netflix_titles |>
  mutate(adj_before_story = str_extract(description, "\\b\\w+(?= story)")) |>
  filter(!is.na(adj_before_story)) |>
  mutate(adj_before_story = tolower(adj_before_story)) |>
  anti_join(stop_words, by = c("adj_before_story" = "word")) |>
  count(adj_before_story, sort = TRUE)

head(story_words, 10)
```

### Most Frequent Words in Descriptions
```{r}
#| fig-alt: "Horizontal bar chart of the 15 most common non-stop words in Netflix descriptions, showing frequency for each word. Words like 'love', 'life', and 'family' appear prominently."
#| eval: true
data("stop_words")

top_words <- netflix_titles |>
  unnest_tokens(word, description) |>
  anti_join(stop_words, by = "word") |>
  count(word, sort = TRUE) |>
  slice_head(n = 15)

ggplot(top_words, aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Most Frequent Words in Netflix Descriptions",
    x = "Word",
    y = "Frequency"
  ) +
  theme_minimal()
```

