---
title: "Project 3: Is Silver More Volatile Than Gold (August–October)?"
author: "Josh Goldhaber"
format:
  html:
    code-fold: true
    theme: flatly
---

# Introduction

In this project, I investigate whether there is a fundamental difference in the forces that drive prices of gold and silver.

Gold and silver are both precious metals often used as investment hedges against inflation and market risk, but their price behaviors differ: gold functions more as a monetary store of value, while silver has heavy industrial demand. Silver has historically traded with higher volatiliy, Although, in recent months, there has been momentum in both metals due to macroeconomic uncertainty.

My analysis aims to determine if there remains a fundamental difference in the factors influencing their price volatility. My permutation test aims to reveal whether Silver's increased volatility can be explained by chance, reflecting a true underlying difference in market dynamics.

Using daily price data from Yahoo Finance (tickers `GC=F` for gold and `SI=F` for silver), I compute daily log returns and perform a one-tailed permutation test under the null hypothesis that the two metals have equal volatility.

Because financial returns are often non-normal, a permutation test is a robust, simulation-based way to test this difference without assuming normality.

------------------------------------------------------------------------

## Experiment Setup

We compare the **standard deviations of daily log returns** for gold and silver.

-   **Null Hypothesis (H₀):** Gold and Silver are driven by the same market factors and their difference in volatility can be explained by chance and randomness of markets.

-   **Alternative Hypothesis (H₁):** The underlying market dynamics for Silver lead to greater daily price variability than Gold that can be explained by randomness or chance.

Instead of relying on a theoretical distribution (like a t-test or F-test), we use a **permutation test**, which simulates what differences we’d expect **purely due to chance** under the null hypothesis.

------------------------------------------------------------------------

## Load Packages and Prepare Data

We use `quantmod` to pull daily futures data from Yahoo Finance and `tidyverse` for data wrangling. Log returns are computed using `dailyReturn(..., type = "log")` for stability and symmetry.

```{r setup, message=FALSE, warning=FALSE}
library(quantmod)
library(tidyverse)
library(infer)
library(purrr)

# Download daily futures prices (August 2025 – Present)
getSymbols("GC=F", src = "yahoo", from = "2025-08-01", to = Sys.Date())
getSymbols("SI=F", src = "yahoo", from = "2025-08-01", to = Sys.Date())

# Compute log returns
gold_ret <- dailyReturn(Cl(`GC=F`), type = "log")
silver_ret <- dailyReturn(Cl(`SI=F`), type = "log")

# Combine and reshape to long format
returns_long <- tibble(
  date = index(gold_ret),
  Gold = as.numeric(gold_ret),
  Silver = as.numeric(silver_ret)
) |>
  drop_na() |>
  pivot_longer(cols = c(Gold, Silver), names_to = "Metal", values_to = "Return")
```

------------------------------------------------------------------------

## Plot Daily Returns

This plot shows the raw time series of **daily log returns** for gold and silver.\
It helps the reader visually compare **volatility over time** — wider fluctuations suggest higher variance.

```{r}

ggplot(returns_long, aes(x = date, y = Return, color = Metal)) +
  geom_line(alpha = 0.6) +
  labs(
    title = "Daily Log Returns for Gold and Silver (August–Present)",
    x = "Date", y = "Daily Log Return",
    color = "Metal"
  ) +
  theme_minimal()
```

**Analysis:**\
We observe that silver (in blue) exhibits more frequent and larger daily swings compared to gold.\
This visual pattern suggests that silver may indeed have higher volatility — which we now test more formally.

------------------------------------------------------------------------

## Define Permutation Function

This function performs **one simulation** of the null hypothesis by randomly shuffling the `Metal` labels.\
This breaks the link between returns and metal type, simulating a world where there is no real difference in volatility.

```{r simulate-diff}
# This function simulates one permutation of the null hypothesis
simulate_diff_sd <- function(df) {
  df |>
    mutate(Metal = sample(Metal)) |>
    group_by(Metal) |>
    summarise(sd = sd(Return), .groups = "drop") |>
    summarise(diff_sd = diff(sd)) |>
    pull(diff_sd)
}
```

------------------------------------------------------------------------

## Observed Difference in SD

We compute the **actual observed difference** in standard deviations — this is our test statistic.\
A positive value means silver is more volatile than gold.

```{r observed-diff}
obs_diff <- returns_long |>
  group_by(Metal) |>
  summarise(sd = sd(Return), .groups = "drop") |>
  summarise(diff_sd = diff(sd)) |>
  pull(diff_sd)

obs_diff
```

**Interpretation:**\
This number captures how much larger silver's return variability is, compared to gold’s.\
We’ll compare this observed difference to the null distribution next.

------------------------------------------------------------------------

## Simulate Null Distribution

We repeat the permutation process **1000 times** using `map_dbl()` to create a distribution of SD differences expected under the null hypothesis (i.e., if metals were equally volatile). The map\_() function allows us to repeat this process smoothly and efficiently.

```{r simulate-null}
set.seed(47)

# Perform 1000 permutations under the null hypothesis
sim_diffs <- map_dbl(1:1000, ~simulate_diff_sd(returns_long))

# Store in tibble for plotting
sim_df <- tibble(diff_sd = sim_diffs)
```

------------------------------------------------------------------------

## Visualize Permutation Distribution

This histogram shows the **null distribution** of SD differences.\
The red dashed line represents the actual observed difference.\
If it's far to the right, that provides evidence that silver is truly more volatile.

```{r plot-null}
ggplot(sim_df, aes(x = diff_sd)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  geom_vline(xintercept = obs_diff, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Permutation Distribution of Difference in Volatility (Silver − Gold)",
    subtitle = "Red dashed line shows the observed difference",
    x = "Difference in SD under Null (permuted)",
    y = "Count"
  ) +
  theme_minimal()
```

**Interpretation:**\
The observed value lies far in the upper tail of the distribution — suggesting that such a large difference would be **very unlikely by chance** if the metals were equally volatile.

------------------------------------------------------------------------

## Calculate One-Tailed p-value

We calculate the **proportion of permuted SD differences** that are greater than or equal to the observed one.\
This gives a **right-tailed p-value** for the hypothesis that silver is more volatile than gold.

```{r p-value}
# One-tailed test: is Silver more volatile than Gold?
p_val <- mean(sim_diffs >= obs_diff)

cat("Observed difference in SD (Silver − Gold):", round(obs_diff, 5), "\n")
cat("Approximate one-tailed permutation p-value:",
    format(p_val, scientific = TRUE, digits = 10), "\n")
```

**Interpretation:**\
A very small p-value implies that the observed volatility difference is unlikely under the null —\
we have statistical evidence to **reject the null hypothesis** and conclude that **silver is more volatile**.

## Simulation: Type I Error Rate of the Permutation Test

We now simulate a world where **gold and silver have the same true volatility**, and check how often our permutation test **falsely rejects the null**.

This helps us assess how well the test controls the Type I error rate.

```{r simulate-type1}
simulate_type1_error <- function(n = 30, reps = 20, perms = 100, seed = 47) {
  set.seed(seed)  # ensures reproducibility

  p_vals <- numeric(reps)

  for (i in 1:reps) {
    group1 <- rnorm(n)
    group2 <- rnorm(n)

    df <- tibble(
      Return = c(group1, group2),
      Metal = rep(c("Gold", "Silver"), each = n)
    )

    obs_diff <- df |>
      group_by(Metal) |>
      summarise(sd = sd(Return), .groups = "drop") |>
      summarise(diff_sd = diff(sd)) |>
      pull(diff_sd)

    sim_diffs <- replicate(perms, {
      df |>
        mutate(Metal = sample(Metal)) |>
        group_by(Metal) |>
        summarise(sd = sd(Return), .groups = "drop") |>
        summarise(diff_sd = diff(sd)) |>
        pull(diff_sd)
    })

    p_vals[i] <- mean(sim_diffs >= obs_diff)
  }

  mean(p_vals < 0.05)
}

simulate_type1_error(n = 30, reps = 20, perms = 100)

```

**Interpretation:**\
This number tells us how often the permutation test **incorrectly rejects H₀** when there is no true difference.

While it differs depending on the seed, 47 gives us .05. The test is behaving well, giving us a Type I Error only .05 of the time If much higher or lower, it may be too liberal or conservative under certain conditions.

## Final Conclusion

This simulation-based test supports the claim that silver exhibits **greater daily price variability** than gold from August to October 2025.\
Given the industrial use of silver and its smaller market size, this result is consistent with economic intuition: silver’s price tends to react more sharply to both macroeconomic shocks and commodity supply disruptions.

The one-tailed permutation test not only confirms this difference but does so **without relying on normality assumptions**, making it a robust method for comparing volatility in real-world financial data.