---
title: "SQL Wrangling"
author: "Josh Goldhaber"
format:
  html:
    code-fold: true
    echo: true
    warning: false
    message: false
    fig-cap-location: bottom
editor: visual
---


# In this project I use three Midwestern policing datasets from the Stanford Open Policing Project (Pierson et al. 2020): Illinois State Patrol, Madison Police Department (WI), and Fort Wayne Police Department (IN). All data wrangling is conducted entirely in SQL. I focus on two central questions: (1) how search rates vary across race and gender, and (2) whether “hit rates” (the percent of searches that find contraband) differ across agencies. The goal is to compare policing practices across these jurisdictions and identify meaningful differences in how traffic stops escalate into searches.
```{r}

library(tidyverse)
library(DBI)
library(ggplot2)


con_traffic <- DBI::dbConnect(

  RMariaDB::MariaDB(),

  dbname = "traffic",

  host = Sys.getenv("TRAFFIC_HOST"),

  user = Sys.getenv("TRAFFIC_USER"),

  password = Sys.getenv("TRAFFIC_PWD")

)

```

```{sql connection=con_traffic}
SHOW TABLES;

```

```{sql connection=con_traffic}
DESCRIBE il_chicago_2023_01_26;


```

```{sql connection=con_traffic}
DESCRIBE wi_madison_2023_01_26;
```

```{sql connection=con_traffic}
DESCRIBE in_fort_wayne_2020_04_01;
```

```{sql connection=con_traffic}
DESCRIBE ky_louisville_2023_01_26;
```



```{sql connection=con_traffic}

# I tried to use a regex to retrieve the state from each table, but the server does not allow SQL regex expressions, so I reserached an interesting workaround: SUBSTRING_INDEX
SELECT 
  table_name,
  SUBSTRING_INDEX(table_name, '_', 1) AS state
FROM information_schema.tables
WHERE table_schema = DATABASE();
```

-----------

-----

```{sql connection=con_traffic, output.var="three_city"}
SELECT *
FROM (
    -- Chicago
    SELECT
        'Chicago' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        citation_issued,
        NULL AS warning_issued
    FROM il_chicago_2023_01_26
    WHERE subject_race IS NOT NULL

    UNION ALL

    -- Madison
    SELECT
        'Madison' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        citation_issued,
        warning_issued
    FROM wi_madison_2023_01_26
    WHERE subject_race IS NOT NULL

    UNION ALL

    -- Louisville
    SELECT
        'Louisville' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        citation_issued,
        warning_issued
    FROM ky_louisville_2023_01_26
    WHERE subject_race IS NOT NULL
) AS t;

```

---
```{sql connection=con_traffic}
CREATE TEMPORARY TABLE three_city AS
SELECT
    'Chicago' AS city,
    subject_race,
    subject_sex,
    date,
    time,
    -- Chicago fix: derive citation indicator
    CASE 
        WHEN outcome = 'Citation Issued' THEN 1
        ELSE 0
    END AS citation_issued,
    NULL AS warning_issued
FROM il_chicago_2023_01_26
WHERE subject_race IS NOT NULL

UNION ALL

SELECT
    'Madison' AS city,
    subject_race,
    subject_sex,
    date,
    time,
    citation_issued,
    warning_issued
FROM wi_madison_2023_01_26
WHERE subject_race IS NOT NULL

UNION ALL

SELECT
    'Louisville' AS city,
    subject_race,
    subject_sex,
    date,
    time,
    citation_issued,
    warning_issued
FROM ky_louisville_2023_01_26
WHERE subject_race IS NOT NULL;

```


```{r}
race_city_summary
```



```{r}

ggplot(race_city_summary,
       aes(x = subject_race,
           y = citation_rate,
           fill = city)) +
  geom_col(position = "dodge") +
  labs(
    title = "Citation rate by driver race across three cities",
    x = "Driver race",
    y = "Share of stops with a citation",
    fill = "City"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

