---
title: "Traffic Stops with SQL"
author: "Josh Goldhaber"
format:
  html:
    code-fold: true
    echo: true
    warning: false
    message: false
    fig-cap-location: bottom
editor: visual
---

# Introduction

This project analyzes traffic-stop data from three agencies in the Stanford Open Policing Project (Pierson et al. 2020): New Orleans Police Department (LA), Madison Police Department (WI), and Louisville Metro Police Department (KY). All data wrangling is done in SQL. R is used only to summarize the SQL output and produce figures.

I focus on two questions:

1.  How do citation and warning rates vary across race within each city?
2.  How do these patterns differ across jurisdictions?

Because the three agencies store outcomes differently, the key step is to harmonize the outcome definitions (especially for New Orleans) and then stack the agencies into one consistent dataset that can be grouped and compared.

```{r}
library(tidyverse)
library(DBI)
library(ggplot2)
library(dplyr)

con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

## Inspecting the database structure

Before building the combined dataset, I confirm table names available in the database. I also extract the state prefix from each table name using `SUBSTRING_INDEX`, which makes it easier to locate relevant datasets quickly.

```{sql connection=con_traffic}
SELECT 
  table_name,
  SUBSTRING_INDEX(table_name, '_', 1) AS state
FROM information_schema.tables
WHERE table_schema = DATABASE()
ORDER BY state, table_name;
```

From this list, I focus on three city-level datasets: `la_new_orleans_2020_04_01`, `wi_madison_2023_01_26`, and `ky_louisville_2023_01_26`.

## Building a three-city dataset in SQL

The block below creates one unified dataset by stacking the three city tables using `UNION ALL` and adding a `city` label. New Orleans stores `outcome` as text, so I use `CASE WHEN` to create comparable binary indicators for `citation_issued` and `warning_issued`. Madison and Louisville already contain numeric flags with those names, so I select them directly. I filter to rows with non-missing `subject_race` because race is essential for the comparisons that follow.

```{sql, connection=con_traffic, output.var="race_city_summary"}
SELECT
  city,
  subject_race,
  COUNT(*) AS stops,
  AVG(citation_issued) AS citation_rate,
  AVG(warning_issued) AS warning_rate
FROM (
    -- New Orleans: recode text outcome into harmonized binary indicators
    SELECT
        'New Orleans' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        CASE WHEN outcome = 'citation' THEN 1 ELSE 0 END AS citation_issued,
        CASE WHEN outcome = 'warning' THEN 1 ELSE 0 END AS warning_issued
    FROM la_new_orleans_2020_04_01
    WHERE subject_race IS NOT NULL

    UNION ALL

    -- Madison: indicators already provided as numeric flags
    SELECT
        'Madison' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        citation_issued,
        warning_issued
    FROM wi_madison_2023_01_26
    WHERE subject_race IS NOT NULL

    UNION ALL

    -- Louisville: indicators already provided as numeric flags
    SELECT
        'Louisville' AS city,
        subject_race,
        subject_sex,
        date,
        time,
        citation_issued,
        warning_issued
    FROM ky_louisville_2023_01_26
    WHERE subject_race IS NOT NULL
) AS three_city
GROUP BY city, subject_race
ORDER BY city, subject_race;

```

# Descriptive summaries

To compare outcomes across race and city, I compute (1) the number of stops and (2) the fraction of stops resulting in citations and warnings. Since `citation_issued` and `warning_issued` are 0/1 indicators, their means are directly interpretable as rates.

This table provides the baseline comparisons. Differences across cities reflect jurisdiction-level enforcement patterns, while differences across race within a city suggest potential disparities in how stops are resolved.

# Visualizing citation and warning rates

I present two bar charts to compare rates across cities and racial groups. Visuals help highlight differences that can be hard to interpret from a table alone, especially when comparing multiple categories simultaneously.

## Figure 1: Citation rates by race and city

```{r}
#| fig-alt: "Bar chart showing citation rates by race across three cities: New Orleans, Madison, and Louisville. The x-axis represents the cities, and the y-axis shows the citation rates as a proportion of stops. Bars are grouped by race, with each race category represented by a different color. The chart highlights differences in citation rates both across cities and within racial groups, with New Orleans generally having lower citation rates compared to Madison and Louisville."

ggplot(race_city_summary, aes(x = city, y = citation_rate, fill = subject_race)) +
  geom_col(position = "dodge") +
  labs(
    title = "Citation Rates by Race Across Three Cities",
    x = "Race",
    y = "Citation Rate (Proportion of Stops)",
    fill = "City"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 1 shows substantial cross-city differences in how often traffic stops result in citations. Overall,New Orleans has consistently lower citation rates across race categories (roughly in the 20%–40% range), while Madison and Louisville are much higher for most groups (often around 70%–85%), suggesting different enforcement practices or recording conventions across jurisdictions.

Within-city racial patterns also vary. In Madison, citation rates are relatively similar across race categories (a fairly “flat” profile), whereas Louisville shows more dispersion, including notably low citation rates for the “other” category and extremely high rates for “unknown.” The “unknown” race category should be interpreted cautiously, since it can reflect missing/uncoded records and may involve smaller sample sizes; a useful follow-up is to compare the stop counts (`stops`) for each race–city group to see whether these extremes are driven by low-N categories.

## Figure 2: Warning rates by race and city

```{r}

#| fig-alt: there  
ggplot(race_city_summary, aes(x = city, y = warning_rate, fill = subject_race)) +
  geom_col(position = "dodge") +
  labs(
    title = "Warning Rates by Race Across Three Cities",
    x = "Race",
    y = "Warning Rate (Proportion of Stops)",
    fill = "City"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 2 shows that warning rates vary more across cities than across racial groups. Madison has the highest warning rates across most races, generally around the high-20% to low-30% range. New Orleans is typically slightly lower but still comparable, while Louisville tends to have the lowest warning rates for most categories. Within each city, differences by race are generally modest. The main exception is Louisville’s “other” category, which appears extremely high (around 80%). This kind of spike is often driven by small sample sizes or differences in how race categories are coded, so it should be interpreted cautiously without checking the number of stops in that group.

Finally, the “unknown” category appears only for New Orleans in this figure, suggesting that missing or unknown race values are coded differently across agencies (or are not present after filtering) and therefore may not be directly comparable across cities.

# Conclusion

Using three city-level datasets from the Stanford Open Policing Project, I constructed a unified dataset using SQL only. The key steps were harmonizing New Orleans outcomes with `CASE WHEN`, stacking the three agencies using `UNION ALL`, and filtering to non-missing race values using `WHERE`. After building the combined table, I summarized stop counts and outcome rates by race and city and visualized citation and warning rates to compare patterns across jurisdictions.

Overall, the results show clear differences in outcome rates across cities and meaningful variation across racial groups within cities. This illustrates how SQL-based harmonization across agencies can reveal cross-city enforcement differences that are difficult to see when looking at one dataset at a time.

# References

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, Phoebe Barghouty, Cheryl Phillips, Ravi Shroff, and Sharad Goel. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops across the United States.” Nature Human Behaviour 4: 736–745. Data from the Stanford Open Policing Project.